---
title: "Sample design using the U.S. Census PDB"
author: "Stas Kolenikov"
date: "21 July 2021"
# output: html_document
output: 
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css"]
    nature:
      highlightStyle: solarized-dark
      highlightLines: true
      countIncrementalSlides: false
  ioslides_presentation:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=5.5, fig.width=8)
#' Format a table for HTML output with striped rows if HTML, or simply display if interactive
#' @param   x an input table
#' @return  a formatted table to be displayed if necessary
#' @export
maybe_kable <- function(x, bg='black') {
  # x is likely a data frame or a tibble
  if (!is.data.frame(x)) return(x)
  else {
    if (interactive()) return(x)
    else {
      # make it a nice kable
      x %>% 
        knitr::kable(format.args = list(big.mark=" ")) %>% 
        kableExtra::row_spec(0,color='white',background="#48A9C5") %>% 
        row_spec(1:nrow(x), color=ifelse(bg=='black', '#d0d3d4', 'black') ) -> tx # header colors
      if (nrow(x)>1) { # alternating rows 
        if (bg=='white') {
          tx %>% row_spec(seq(from=2,to=nrow(x),by=2), background = '#dfd1a7') -> tx
        } else if (bg=='black') {
          tx %>% row_spec(seq(from=2,to=nrow(x),by=2), background = '#789d4a') -> tx 
        }
      }
      return(tx)
    }
  }
}

```

## Show of hands

1. Name!
2. Where you are from (location and affiliation).
3. The most interesting thing you've learned so far in SPSS.
4. R? Stata? SAS? SPSS? Python? MATLAB?

---

## Outline

1. The U.S. Census Planning Database, U.S. Census geography, race/ethnicity
2. Survey sampling design target
3. Stas' initial attempt
4. Actual workshop -- challenge to improve upon Stas' work!

---

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
libs <- c('tidyverse', 'here', 'knitr', # 'vtable',
          'xaringanthemer', 'kableExtra')
for( l in libs) {
  library(l, character.only = TRUE)
}
```

The best way to proceed with this workshop is:

1. Create a new RStudio project for this workshop in a new folder.
1. Download the .Rmd file from Canvas.
1. Download the .csv file from Canvas into PDB subfolder.
1. Open the .Rmd file and "Run all chunks" from RStudio menu.
1. `install.packages("library_name")` whatever you may  be missing.

In Stata, you should be able to read the data file as:

```
import delimited using PDB/pdb2021trv3_ct.csv, clear
```

```{r technical_stuff, echo=FALSE}
style_solarized_dark()
style_xaringan(
  background_color = '#000000',
  text_color       = '#b7c9d3',
  header_color     = '#DA291C',
  title_slide_text_color = '#DA291C',
  title_slide_background_color = '#DFD1A7',
  title_slide_background_image = 'AbtLogo2.png',
  title_slide_background_size	= '200px 100px',
  title_slide_background_position	= 'bottom 10px left 20px',
  inverse_text_color = '#000000',
  inverse_header_color = '#7566A0',
  inverse_background_color = '#DFD1A7',
  inverse_text_shadow = TRUE,
  background_image = 'AbtLogo2.png',
  background_size	= '90px 45px',
  background_position	= 'bottom 10px left 10px',
  link_color       = '#7566A0',
  code_inline_color = '#C3C6A8',
  code_highlight_color = '#E87722'
)
```
---

class: inverse

# The U.S. Census Bureau and its data

---

## Planning Databases

https://www.census.gov/topics/research/guidance/planning-databases.2021.html

---

## PDB data

```{r pdb, message=FALSE, warning=FALSE}
if (file.exists(here('PDB','pdb2021trv3_ct.csv'))) {
  PDB_CT <- read_csv(here('PDB','pdb2021trv3_ct.csv'))
} else {
  PDB_US <- read_csv(here('PDB','pdb2021trv3_us.csv'))
}
```

*Note to students*: the above code uses `library(here)` to create 
internalized relative links to files within an R/RStudio project.
The reference `here('PDB','pdb2021trv3_ct.csv')` will point to the file
in the directory `PDB`, i.e., will create a full reference 

> `[project root folder]/PDB/pdb2021trv3_ct.csv`

On my computer, this reference is `r here('PDB','pdb2021trv3_ct.csv')`. 
On your computer, you would need to save the `.Rmd` file with this presentation 
and either

- start an RStudio project in that folder, or
- create an empty `.here` file in that same directory so that `library(here)` will pick that
  up as the starting location.


---

## US Census Tracts

* Tract $\subset$ county $\subset$ state

* Tract population: about 4000

https://www2.census.gov/geo/pdfs/education/CensusTracts.pdf

---

## Example tracts

University of Michigan:

- ACS profile ([MCDC](http://mcdc.missouri.edu/applications/acs/profiles/report.php?period=5&year=2018&g=14000US26161400200|05000US26161|04000US26|01000US))
- Tiger boundaries: https://tigerweb.geo.census.gov/tigerweb/, 
search for 500 S STATE ST, ANN ARBOR, MI, 48109

Stas' residence:

- ACS profile ([MCDC](http://mcdc.missouri.edu/applications/acs/profiles/report.php?period=5&year=2018&g=14000US29019001103|05000US29019|04000US29|01000US))
- Tiger boundaries: https://tigerweb.geo.census.gov/tigerweb/, 
search for CT 11.08 COLUMBIA, MO, 65203

---

class: inverse

# Sample design task

---

## Sample design target

We need to create a sample of adults in the state of 
[https://en.wikipedia.org/wiki/Connecticut](Connecticut),
with the target of `r (target_n <- 2500)`, 
and oversample targets for racial/ethnic minorities:

* Black/African American: `r (target_black <- 500)`
* Hispanic: `r (target_hisp <- 500)`

Simplifications:
* disregard household size distributions between race/ethnicity groups
* disregard the age distributions between race/ethnicity groups

---

## Connecticut

```{r CT, message=FALSE}
if (!exists("PDB_CT")) {
  PDB_US %>% filter(State=='09', !is.na(Tot_Population_ACS_15_19), 
               Tot_Population_ACS_15_19 > 0 ) -> PDB_CT 
  # export
  write_csv(PDB_CT, here('PDB','pdb2021trv3_ct.csv'))
}
PDB_CT %>% 
  group_by(State, State_name) %>% 
  summarize(tracts=n(), 
    adult_pop = sum( Pop_18_24_ACS_15_19 + Pop_25_44_ACS_15_19 + 
                   Pop_45_64_ACS_15_19 + Pop_65plus_ACS_15_19) ) %>% 
  maybe_kable()
```

---

class: inverse

# Stas' first steps

---

## Solution: stratified design

Create several strata and vary sampling rates between them 
to achieve the target sample sizes.

```
PDB_CT %>% mutate( 
  pct_NH_black_alone = NH_Blk_alone_ACS_15_19 / Tot_Population_ACS_15_19,
  pct_hisp           = Hispanic_ACS_15_19 / Tot_Population_ACS_15_19,
  pct_minority       = pct_NH_black_alone + pct_hisp
) -> PDB_CT
ggplot(data=PDB_CT) + 
  geom_histogram(aes(x=pct_minority), color='skyblue')
```

---

## Solution: stratified design

Create several strata and vary sampling rates between them 
to achieve the target sample sizes.

```{r pct_minority, message=FALSE, warning=FALSE, echo=FALSE}
PDB_CT %>% mutate( 
  pct_NH_black_alone = NH_Blk_alone_ACS_15_19 / Tot_Population_ACS_15_19,
  pct_hisp           = Hispanic_ACS_15_19 / Tot_Population_ACS_15_19,
  pct_minority       = pct_NH_black_alone + pct_hisp
) -> PDB_CT
ggplot(data=PDB_CT %>% mutate(pct_minority = pct_minority*100)) + 
  geom_histogram(aes(x=pct_minority), color='#48A9C5') +
  xlab('% non-white') + ylab('# of tracts') + theme_light()
```

---

## Beware of nonresponse!

```
ggplot(data=(PDB_CT) +
  geom_point(aes(x=pct_minority,y=Low_Response_Score,
                 size=Tot_Population_ACS_15_19) )
```

```{r minority_vs_LRS, warning=FALSE, echo=FALSE}
ggplot(data=(PDB_CT %>% rename(`Tract Pop`=Tot_Population_ACS_15_19) %>% mutate(pct_minority=pct_minority*100))) +
  geom_point(aes(x=pct_minority,y=Low_Response_Score,
                 size=`Tract Pop`),
                 fill='#48A9C5', colour='black', shape=21) +
  xlab('% non-white') + ylab('Low Response Score') +
  theme_light()
```

---

## Overall anticipated response rate

```{r RR}
PDB_CT %>% summarize( 
  LRS = weighted.mean(x=Low_Response_Score,
                      w=Tot_Population_ACS_15_19, 
                      na.rm=TRUE) ) %>%
  select(LRS) %>% unlist() -> LRS
```

The overall, population weighted low response score is `r LRS`.

Thus for the target sample size of 2500, one needs to field about 
`r (n_field <- ceiling(target_n/(1-LRS/100)))` cases.

---

## Two-strata solution: high vs. low minority tracts

```{r two_strata}
PDB_CT %>% mutate(strata2 = if_else(pct_minority > 0.5, 1, 2) ) %>%
  group_by(strata2) %>%
  summarize(
    tract        = n(),
    min_minority = min(pct_minority),
    max_minority = max(pct_minority),
    pop   = sum(Tot_Population_ACS_15_19),
    black = sum(NH_Blk_alone_ACS_15_19),
    hisp  = sum(Hispanic_ACS_15_19),
    RR    = 1 - weighted.mean(x=Low_Response_Score,
                              w=Tot_Population_ACS_15_19, 
                              na.rm=TRUE)/100  
) -> CT_strata2
```

---

## Two-strata solution: high vs. low minority tracts

```
CT_strata2 %>% kable()
```

```{r CT_strata2, echo=FALSE}
CT_strata2 %>% maybe_kable()
```

---

## Trial-and-error allocation

Compute anticipated number of Black/AA interviews; number of Hispanic interviews;
adjust inputs until the results are acceptable 

```{r trial_error2}
CT_strata2 %>% mutate(
  n_field = case_when(strata2 == 1 ~ 2100, strata2 == 2 ~ 1100),
  n_total = floor(n_field * RR),
  n_black = floor(n_field * RR * black / pop),
  n_hisp  = floor(n_field * RR * hisp / pop),
  sampling_rate = n_field/pop*1e3
) %>% select(strata2, sampling_rate, starts_with('n_')) -> CT_strata2_completes
```

```{r trial_error2_print, echo=FALSE}
maybe_kable(CT_strata2_completes)
```

Overall sample size: `r sum(CT_strata2_completes$n_total)` vs. `r target_n`, 
Black AA race and Hispanic ethnicity oversamples of 
`r sum(CT_strata2_completes$n_black)` and `r sum(CT_strata2_completes$n_hisp)`
(vs. target `r target_hisp` each).


---

## Trial-and-error allocation

Compute anticipated number of Black/AA interviews; number of Hispanic interviews;
adjust inputs until the results are acceptable 
(overall sample size `r target_n`, 
Black AA race/Hispanic ethnicity oversamples of `r target_hisp` each).

```{r trial_error2a, echo=FALSE}
CT_strata2 %>% mutate(
  n_field = case_when(
    strata2 == 1 ~ 2115,
    strata2 == 2 ~ 1230
  ),
  n_total = floor(n_field * RR),
  n_black = floor(n_field * RR * black / pop),
  n_hisp  = floor(n_field * RR * hisp / pop),
  sampling_rate = n_field/pop*1e3
) %>% select(strata2, sampling_rate, starts_with('n_')) -> CT_strata2_completes
```

```
bind_rows( CT_strata2_completes %>% mutate(strata2=as.character(strata2)),
           CT_strata2_completes %>% select(starts_with("n_")) %>% 
           summarize_all(sum) %>% mutate(strata2='Total')) %>% kable()
```

```{r CT_strata2_results, echo=FALSE}
bind_rows( CT_strata2_completes %>% mutate(strata2=as.character(strata2)),
           CT_strata2_completes %>% select(starts_with("n_")) %>% 
           summarize_all(sum) %>% mutate(strata2='Total') 
  ) %>% maybe_kable() 
```

---

## Simple weights

```{r weights}
CT_strata2_completes %>% 
  full_join( CT_strata2 %>% select(strata2, pop), by='strata2') %>%
  mutate(weight=pop/n_total) %>% 
  select(strata2, n_field, n_total, pop, weight) -> CT_strata2_weights
CT_strata2_weights %>% maybe_kable()
```

---

## Unequal weighting design effect

Unequal weighting design effect $1+{\rm CV}^2 \equiv 1 + L_{Kish}$ for this design is:
```{r deff}
(CT_strata2_weights %>% 
  summarise( n_wgt = sum(n_total*weight), 
             n_wgt2 = sum(n_total*weight*weight),
             n = sum(n_total) ) %>% 
  mutate(UWE_DEFF = n_wgt2 * n / (n_wgt*n_wgt) ) %>% 
  select(UWE_DEFF) %>% unlist() -> UWE_DEFF2)
```

---

class: inverse

# Can you do better??


---

## Better solutions?

* Better choice of the threshold in a two-strata solution?
* Three strata? 
    - two thresholds of minorities, combined?
    - separate thresholds for Black/African Americans vs. Hispanics?
* Four strata?
* Minimize design effect?
* Account for response rates at the tract level?

---

## Your turn now!

**Exercise 1**

* Breakout Zoom rooms, groups of ~4
* Create a _better_ design:
    - the above one had too many Hispanics relative to the target 
      (and relative to SRS; hence losses of efficiency)

Stas' best design has DEFF of about 1.23, but it involved heavy-handed numeric optimization
with tons of fiddling with optimization parameters.

---

## Further refinements

* Adult vs. total population
* Language barriers (speak English less than very well)
    - partially incorporated in the Low Response Score
    - limits the covered population
* Residential households (vs. group quarters)
* Vacant housing units
* Lower response rates to non-federal surveys

---

## Exercise 2: adjust response rate

Based on your prior experience / existing literature on the surveys of this kind,
you expect the overall response rate to be 15%.

Adjust the parameters of your sampling design that you have produced so far
in a way that the state-wide response rate on a SRS will be 15%.
The `Low_Response_Score` variable will be informative on the relative
performance of the different census tracts, but it would have to be modified 
so that the anticipated response rates hover around the above target figure.
This is an open-ended task: you would need to come up with a (mostly) justifiable
approach to the task. 
(A lot of survey statistics consists of inventing or adjusting things!)

---

## Exercise 3: cost-optimal design

Suppose that you want to enroll more Hispanic respondents to the survey
by providing materials in Spanish. Let us use the following assumptions:

- the cost of the mailing increases by a factor of 2;
- the response rate of Hispanic respondents who receive materials in Spanish 
  increases by 50% (e.g. from 10% to 15%);
- the Spanish mail materials are only sent in the stratum with the highest
  concentration of Hispanic population;
- the available budget allows for the mailing of 24,000 "basic" mail materials
  in English only.

Objective: adjust the sampling design to maximize the overall effective sample size
while maintaining the subpopulation sample sizes of 500 Black/African American respondents
and 500 Hispanic respondents.

Does offering the instrument in Spanish really save money?

---

## R Markdown

This is an R Markdown `library(xaringan)` presentation. 
Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. 
For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

R version: `r R.version.string`.

Package versions:

```{r package_versions, echo=FALSE}
pvers <- ''
for( p in libs ) {
  pvers <- paste0(pvers, 
      '- library(',p, '):  version ', as.character(packageVersion(p)), 
      '\n')
}
```

`r pvers`

The color scheme used is the corporate scheme of Abt Associates.
